---
title: "CodeBook"
author: "Julian Posada"
date: "2023-02-13"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

### This is de CodeBook for the final project: Peer-graded Assignment: Getting and Cleaning Data Course Project
### Johns Hopkins Data Science Specialization via Coursera 

One of the most exciting areas in all of data science right now is wearable computing.
Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users. 
The data linked to from the course website represent data collected from the accelerometers from the Samsung Galaxy S smartphone. A full description is available at the site where the data was obtained:

[archive.ics.uci.edu](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones)

**Here are the data for the project:**

 [cloudfront.net](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip )

You should create one R script called run_analysis.R that does the following. 

Merges the training and the test sets to create one data set.

Extracts only the measurements on the mean and standard deviation for each measurement. 

Uses descriptive activity names to name the activities in the data set

Appropriately labels the data set with descriptive variable names. 

From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.


# My project: 

the explanation of the variables can be found at [codebook.docx](https://github.com/JulianPosadaDenHaag/Johns_Hopkins_Coursera_Data_Science/blob/7ef7711c0324055b5f5cc74f574139d42736d12d/Getting%20and%20Cleaning%20Data%20Course%20Project/codebook.docx)

or at the end of this document with a skimr summary

First I will load the packages I need to perform the getting and cleaning process


```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(readr)
library(skimr)
library(data.table)
library(stringr)
```

Using the library here, I am setting up the correct path
```{r}
path<-     here::here( "Getting and Cleaning Data course Project","UCI HAR Dataset")
```

The next code  takes the features names from "features.txt"

```{r}
features<-data.table::fread(here::here(path,"features.txt"))%>% 
  set_names(c("index", "names")) 
str(features)
```

this select only the "names" in "features.txt" that contains mean or std
```{r}
mean_std<- features %>% filter( grepl("mean|std",features$names))
str(mean_std)
```

Loading the x_train.txt (training data set) into a data.table, and setting the names of all columns to be the right
names (in features). Then only selecting the names that contains mean or std 

```{r}
x_train<- data.table::fread(here::here(path, "train", "x_train.txt"))%>%
  setNames(features$names) %>% select(mean_std$names)
head(as.tibble(x_train))
```


A data.table is created with the values to be transformed into activity_labels

```{r}
y_train<-  data.table::fread(here::here(path, "train","y_train.txt")) %>% 
        set_names("labels")
str(y_train)
```


A data.table is created that contains all te subjects (as character).  An index is created. 
```{r}
subject_train<-  data.table::fread(here::here(path, "train","subject_train.txt"),
                                   colClasses = "character") %>%
        set_names("subject")%>% 
        mutate("index" =row_number())
str(subject_train)
```



The data from y_train, subject_train and activity labels is then merged. 
```{r}
activity_labelsX<-data.table::fread(here::here(path,"activity_labels.txt"))%>% 
         set_names(c("labels", "activity_labels")) %>% 
        merge(y_train, by = "labels") %>% mutate("index" =row_number()) %>% 
        merge(subject_train, by = "index") %>% select(!index & !labels)
str(activity_labelsX)
```

All the columns are merged to created the tidy x_train dataset. 
```{r}
x_train<-bind_cols(activity_labelsX,x_train)
head(as.tibble(x_train))
```


The same process is repeated for the test dataset
```{r}
x_test<- data.table::fread(here::here(path, "test", "X_test.txt"))%>%
        setNames(features$names)%>% select(mean_std$names)

y_test<-  data.table::fread(here::here(path, "test", "y_test.txt")) %>% 
        set_names("labels")

subject_train_y<-  data.table::fread(here::here(path, "test","subject_test.txt"),colClasses = "character") %>%
        set_names("subject")%>% 
        mutate("index" =row_number())

activity_labelsY<-data.table::fread(here::here(path, "activity_labels.txt"))%>% 
        set_names(c("labels", "activity_labels")) %>% 
        merge(y_test, by = "labels") %>% mutate("index" =row_number()) %>% 
        merge(subject_train_y, by = "index")%>% select(!index & !labels)

x_test<-bind_cols(activity_labelsY,x_test) 
head(as.tibble(x_test))
```



Binding all the rows to create a tidy_dataset.  The subject is set to the first column
```{r}
tidy_data<- bind_rows(x_train,x_test) %>% 
        relocate(subject)
head(as.tibble(tidy_data))
```

The tidy data is then grouped and summarized by average 

```{r}
tidy_data_average<- tidy_data %>% 
  group_by( subject, activity_labels) %>% 
  summarise(across(everything(), mean), .groups = "drop") %>% 
  arrange(activity_labels) 
tidy_data_average
skim(tidy_data_average)

```





